### 2021-09-03 revision/ch3bedEle.tex
added phd thesis.

In this work, I present the first stochastic model coupling the erosion and deposition of individual bedload grains to local bed elevation changes.
This model extends the birth-death model approach reviewed in section \ref{sec:birthdeath} to describe the interplay between the bedload flux and the bed elevation within a control volume.
This new model evaluates the effect of collective entrainment on bed elevation change, and it allows for investigations of the residence times of particles buried in the bed subsurface.
The model has two major assumptions: First, bedload erosion and deposition are characterized by probabilities per unit time, or rates \citep{Einstein1950, Ancey2008}; and second, these rates are contingent on the local bed elevation, encoding the property that erosion of sediment is emphasized from regions of exposure, while deposition is emphasized in regions of shelter \citep{Sawai1987, Wong2007}.

@@

Using this method, I simulated 4 transport conditions with 13 different values of $l$ taken across a range from $l=a$ (a single radius) to $l=10a$ (10 radii).
These values include the range exhibited by the available experimental data on bed elevation timeseries \citep{Wong2007,Singh2009,Martin2014}.

@@

A more careful approximate solution to equation (\ref{eq:approxele}) can be obtained by prescribing a phenomenological equation for $\langle n | m \rangle$ into equation (\ref{eq:approxele}) in order to close the equation for $m$ without solving equation (\ref{eq:decomp}).
From numerical simulations we determine that
\begin{equation}
	\langle n | m  \rangle \approx \langle n \rangle \Big( 1 - \frac{2\kappa m}{1-\mu/\sigma}\Big) \label{eq:closure}
\end{equation}
captures the general features of the conditional mean particle activity.
As I show in the supplementary information, introducing this closure equation to (\ref{eq:approxele}), making the Fokker-Planck approximation, and neglecting terms of $O(\kappa^2)$ provides 
\begin{equation} M(m) \approx M_0 e^{-2\kappa m^2}, \label{eq:ou2}\end{equation}
representing a Gaussian distribution with variance $\sigma_m^2 = \frac{1}{4\kappa}$ -- smaller than the former mean field theory by a factor of two and in agreement with the result posited in equation (\ref{eq:active}).
$M_0$ is a normalization constant. 
This closure equation approach shows good coorespondence with numerical solutions of equation (\ref{eq:elemaster}), at least for the flow parameters in table \ref{tab:anceyparams}.

@@

The coupling between bed elevation changes and the erosion and deposition rates develop a strong dependence of the particle activity on $m$. Figure (\ref{fig:condmoms}) displays the mean shift $[\langle n |m \rangle - \langle n \rangle]/\langle n \rangle $ and the variance shift  $[\text{var}(n|m) - \text{var}(n)]/\text{var}(n)$ of the particle activity due to departures of the bed elevation from its mean position.
\begin{figure}[!htbp]
	\includegraphics[width=\linewidth,keepaspectratio]{./figures/ch3/momentsuppression.pdf}
	\caption{The shifts between particle activity moments conditioned on instantaneous elevations and their over-all mean values. Panel (a) indicates the mean particle activity shift versus the bed elevation measured in units of $\sigma_z=l$. This shift displays asymmetric dependence on $m$ at the flow conditions of the \citet{Ancey2008} experiements, and departures of the bedload transport mean can be as much as 60\% when the bed is in a severely degraded state with $z\approx -3l$. The closure equation (\ref{eq:closure}) is plotted in panel (a) Panel (b) demonstrates a more symmetrical variance shift with some dependence on flow conditions displaying shifts of up to 20\% with bed elevations. These results indicate that bedload statistics measurements on short timescales could be severely biased by departures from the mean bed elevation.}
	\label{fig:condmoms}
\end{figure}
Figure (\ref{fig:condmoms}a) demonstrates that the flow conditions in table \ref{tab:anceyparams} support departures of the mean particle activity by as much as 60\% from the overall mean value when the bed is in a degraded state $z\approx -3l$, and the activity can be decreased by 20\% when the bed is in an aggraded state.
The closure model (\ref{eq:closure}) used to derive the approximate bed elevation distribution (\ref{eq:ou2}) is plotted behind the conditional mean profiles in figure (\ref{fig:condmoms}a), where it appears to be crude approximation, as it does not capture the asymmetry.
Nevertheless, figure \ref{fig:var} demonstrates the variance $1/(4\kappa)$ derived from this closure equation is representative of the numerical relationship.
For the parameters of the \citet{Ancey2008} experiments, figure (\ref{fig:condmoms}b) displays a variance shift with bed elevation changes that is less severe than the mean shift but is nevertheless appreciable, with bed elevations changing the magnitude of bedload activity fluctuations by as much as 20\%.
The summary is that bed elevation changes regulate the particle activity moments, with a moment suppression effect when the bed is aggraded, and a moment enhancement effect when the bed is degraded.

@@

Resting times for sediment undergoing burial are obtained from analyzing the return times from above in the time-series of $m$ \citep[e.g.,][]{Redner2007}.
Following \citet{Voepel2013} and \citet{Martin2014}, we concentrate on a particular bed elevation $m'$, and find all time intervals separating deposition events at $m=m'$ from erosion events at $m=m'+1$.
These are the return times from above of the sedimentary bed conditional to the elevation $m'$.
Binning these conditional return times (using logarithmically-spaced bins to reduce computational load) and counting the occurrences in each bin, we obtain an exceedance distribution of return times $t_r$ held conditional to the elevation $m'$: $P(T>t_r|m')$.
Using the marginal probability distribution of bed elevations $P(m)$ (figure \ref{fig:pdfs}(b)), we derive the unconditional exceedance distribution of resting times as a sum over all elevations \citep{Yang1971, Nakagawa1980, Voepel2013, Martin2014}: 
\begin{equation} P(T>t_r) = \sum_{m'} P(m') P(T>t_r|m') .\end{equation}
A representative subset of these results are displayed in figure \ref{fig:cdfs}.
Comparing panels \ref{fig:cdfs}(a) and \ref{fig:cdfs}(c) shows two separate variations with input parameters: first, the distributions vary with the flow conditions, and second, they vary with the standard deviation of bed elevations ($l$).
However, as shown in panels \ref{fig:cdfs}(b) and \ref{fig:cdfs}(d), a characteristic timescale $T_0$ is found to collapse away both variations.
We obtained this $T_0$ heuristically by considering the characteristic speed of bed elevation change.
This is the mean number of grains leaving the bed per unit time is $E$, and the removal of a single grain changes the bed elevation by $z_1$ (\ref{eq:ele}). 
Therefore, bed elevations change with a characteristic speed $v = z_1 E$.
Since the range of elevation deviations is $l$ (figure \ref{fig:var}), the time required for the bed to shift through this characteristic distance is $l/v$, or equivalently
\begin{equation} T_0 = \frac{l}{z_1 E}.\label{eq:time}\end{equation}
When scaling the resting time by this $T_0$, we obtain the collapse shown in figure \ref{fig:cdfs}.
Using the log-likelihood estimation technique described by \citet{Newman2005}, we estimate the scaled resting time non-exceedance distributions decay as a heavy tailed power law with parameter $\alpha = 1.18 \pm 0.32$ for all return times satisfying $T/T_0 > 10^3$.
These distributions are sufficiently heavy tailed to violate the central limit theorem and drive anomalous super-diffusion of bedload, a result which supports the earlier conclusions of \citet{Voepel2013} and \citet{Martin2014}.

@@

Finally, the model allowed resolution of resting times for sediment undergoing burial within the sedimentary bed. These timescales were obtained by analyzing return times from above in the bed elevation timeseries.
This analysis produces heavy-tailed power-law resting times with tail parameters sufficient to drive anomalous diffusion of bedload at long timescales.
The power-law tails were found to collapse across flow conditions using a timescale formed from the mean erosion rate and the active layer depth.

@@

This chapter presented a stochastic model for particle activity and local bed elevations including feedbacks between elevation changes and sediment transport.
This model includes collective entrainment, whereby moving particles tend to destabilize stationary ones.
The model was analyzed using a mixture of numerical and analytical methods, and two key results were presented:
\begin{enumerate}
	\item Resting times for sediment undergoing burial lie on a heavy-tailed power law distributions with tail parameter $\alpha \approx 1.2$;
	\item Collective entrainment generates a statistical regulation effect, whereby bed elevation changes modify the mean and variance of the particle activity by as much as 90\%: this effect vanishes when collective entrainment is absent.
\end{enumerate}
These results imply measurements of bedload transport statistics could be biased at observation timescales smaller than adjustments of the bed elevation timeseries when collective entrainment occurs.
The next step is to generalize the model in this chapter to a multi-cell framework to study channel morphodynamics in a stochastic framework.

>>>>

In this work, I present the first stochastic model coupling the erosion and deposition of individual bedload grains to local bed elevation changes.
This model extends the birth-death model approach reviewed in section \ref{sec:birthdeath} to describe the interplay between the bedload flux and the bed elevation within a control volume.
This new model evaluates the effect of collective entrainment on bed elevation change, and it allows for investigations of the residence times of particles buried under the bed surface.
The model has two major assumptions: First, bedload erosion and deposition are characterized by probabilities per unit time, or rates \citep{Einstein1950, Ancey2008}; and second, these rates are contingent on the local bed elevation, encoding the property that erosion of sediment is emphasized from regions of exposure, while deposition is emphasized in regions of shelter \citep{Sawai1987, Wong2007}.

@@

Using this method, I simulated 4 transport conditions with 13 different values of $l$ taken across a range from $l=a$ (a single radius) to $l=10a$ (10 radii).
These values include the range exhibited by the available experimental data on bed elevation time series \citep{Wong2007,Singh2009,Martin2014}.

@@

A more careful approximate solution to equation (\ref{eq:approxele}) can be obtained by prescribing a phenomenological equation for $\langle n | m \rangle$ into equation (\ref{eq:approxele}) in order to close the equation for $m$ without solving equation (\ref{eq:decomp}).
From numerical simulations we determine that
\begin{equation}
	\langle n | m  \rangle \approx \langle n \rangle \Big( 1 - \frac{2\kappa m}{1-\mu/\sigma}\Big) \label{eq:closure}
\end{equation}
captures the general features of the conditional mean particle activity.
As I show in the supplementary information, introducing this closure equation to (\ref{eq:approxele}), making the Fokker-Planck approximation, and neglecting terms of $O(\kappa^2)$ provides 
\begin{equation} M(m) \approx M_0 e^{-2\kappa m^2}, \label{eq:ou2}\end{equation}
representing a Gaussian distribution with variance $\sigma_m^2 = \frac{1}{4\kappa}$ -- smaller than the former mean field theory by a factor of two and in agreement with the result posited in equation (\ref{eq:active}).
$M_0$ is a normalization constant. 
This closure equation approach shows good correspondence with numerical solutions of equation (\ref{eq:elemaster}), at least for the flow parameters in table \ref{tab:anceyparams}.

@@

The coupling between bed elevation changes and the erosion and deposition rates develop a strong dependence of the particle activity on $m$. Figure (\ref{fig:condmoms}) displays the mean shift $[\langle n |m \rangle - \langle n \rangle]/\langle n \rangle $ and the variance shift  $[\text{var}(n|m) - \text{var}(n)]/\text{var}(n)$ of the particle activity due to departures of the bed elevation from its mean position.
\begin{figure}[!htbp]
	\includegraphics[width=\linewidth,keepaspectratio]{./figures/ch3/momentsuppression.pdf}
	\caption{The shifts between particle activity moments conditioned on instantaneous elevations and their over-all mean values. Panel (a) indicates the mean particle activity shift versus the bed elevation measured in units of $\sigma_z=l$. This shift displays asymmetric dependence on $m$ at the flow conditions of the \citet{Ancey2008} experiements, and departures of the bedload transport mean can be as much as 60\% when the bed is in a severely degraded state with $z\approx -3l$. The closure equation (\ref{eq:closure}) is plotted in panel (a) Panel (b) demonstrates a more symmetrical variance shift with some dependence on flow conditions displaying shifts of up to 20\% with bed elevations. These results indicate that bedload statistics measurements on short time scales could be severely biased by departures from the mean bed elevation.}
	\label{fig:condmoms}
\end{figure}
Figure (\ref{fig:condmoms}a) demonstrates that the flow conditions in table \ref{tab:anceyparams} support departures of the mean particle activity by as much as 60\% from the overall mean value when the bed is in a degraded state $z\approx -3l$, and the activity can be decreased by 20\% when the bed is in an aggraded state.
The closure model (\ref{eq:closure}) used to derive the approximate bed elevation distribution (\ref{eq:ou2}) is plotted behind the conditional mean profiles in figure (\ref{fig:condmoms}a), where it appears as a crude approximation since it does not capture the asymmetry.
Nevertheless, figure \ref{fig:var} demonstrates the variance $1/(4\kappa)$ derived from this closure equation is representative of the numerical relationship.
For the parameters of the \citet{Ancey2008} experiments, figure (\ref{fig:condmoms}b) displays a variance shift with bed elevation changes that is less severe than the mean shift but is nevertheless appreciable, with bed elevations changing the magnitude of bedload activity fluctuations by as much as 20\%.
The summary is that bed elevation changes regulate the particle activity moments, with a moment suppression effect when the bed is aggraded, and a moment enhancement effect when the bed is degraded.

@@

Resting times for sediment undergoing burial are obtained from analyzing the return times from above in the time-series of $m$ \citep[e.g.,][]{Redner2007}.
Following \citet{Voepel2013} and \citet{Martin2014}, we concentrate on a particular bed elevation $m'$, and find all time intervals separating deposition events at $m=m'$ from erosion events at $m=m'+1$.
These are the return times from above of the sedimentary bed conditional to the elevation $m'$.
Binning these conditional return times (using logarithmically-spaced bins to reduce computational load) and counting the occurrences in each bin, we obtain an exceedance distribution of return times $t_r$ held conditional to the elevation $m'$: $P(T>t_r|m')$.
Using the marginal probability distribution of bed elevations $P(m)$ (figure \ref{fig:pdfs}(b)), we derive the unconditional exceedance distribution of resting times as a sum over all elevations \citep{Yang1971, Nakagawa1980, Voepel2013, Martin2014}: 
\begin{equation} P(T>t_r) = \sum_{m'} P(m') P(T>t_r|m') .\end{equation}
Figure \ref{fig:cdfs} displays a representative subset of these results.
Comparing panels \ref{fig:cdfs}(a) and \ref{fig:cdfs}(c) shows two separate variations with input parameters: first, the distributions vary with the flow conditions, and second, they vary with the standard deviation of bed elevations ($l$).
However, as shown in panels \ref{fig:cdfs}(b) and \ref{fig:cdfs}(d), a characteristic time scale $T_0$ is found to collapse away both variations.
We obtained this $T_0$ heuristically by considering the characteristic speed of bed elevation change.
Because the mean number of grains leaving the bed per unit time is $E$ and the removal of a single grain changes the bed elevation by $z_1$ (\ref{eq:ele}), bed elevations change with a characteristic speed $v = z_1 E$.
Since the range of elevation deviations is $l$ (figure \ref{fig:var}), the time required for the bed to shift through this characteristic distance is $l/v$, or equivalently
\begin{equation} T_0 = \frac{l}{z_1 E}.\label{eq:time}\end{equation}
When scaling the resting time by this $T_0$, we obtain the collapse shown in figure \ref{fig:cdfs}.
Using the log-likelihood estimation technique described by \citet{Newman2005}, we estimate the scaled resting time non-exceedance distributions decay as a heavy tailed power law with parameter $\alpha = 1.18 \pm 0.32$ for all return times satisfying $T/T_0 > 10^3$.
These distributions are sufficiently heavy tailed to violate the central limit theorem and drive anomalous super-diffusion of bedload, a result which supports the earlier conclusions of \citet{Voepel2013} and \citet{Martin2014}.

@@

Finally, the model allowed resolution of resting times for sediment undergoing burial within the sedimentary bed. These time scales were obtained by analyzing return times from above in the bed elevation time series.
This analysis produces heavy-tailed power-law resting times with tail parameters sufficient to drive anomalous diffusion of bedload at long time scales.
The power-law tails were found to collapse across flow conditions using a time scale formed from the mean erosion rate and the active layer depth.

@@

This chapter presented a stochastic model for particle activity and local bed elevations including feedbacks between elevation changes and sediment transport.
This model includes collective entrainment, whereby moving particles tend to destabilize stationary ones.
The model was analyzed using a mixture of numerical and analytical methods, and two key results were presented:
\begin{enumerate}
	\item Resting times for sediment undergoing burial lie on a heavy-tailed power law distributions with tail parameter $\alpha \approx 1.2$;
	\item Collective entrainment generates a statistical regulation effect, whereby bed elevation changes modify the mean and variance of the particle activity by as much as 90\%: this effect vanishes when collective entrainment is absent.
\end{enumerate}
These results imply measurements of bedload transport statistics could be biased at observation time scales smaller than adjustments of the bed elevation time series when collective entrainment occurs.
The next step is to generalize the model in this chapter to a multi-cell framework to study channel morphodynamics in a stochastic framework.
